HMM

### 2015-11-06 paper3.Rnw
fixed minor stuff in Rnw

\documentclass{article}
\usepackage{natbib}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\bibliographystyle{chicago}
\title{ HMM}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

\keywords{abc}

@@

Given a sequence of animal movements observed via direct observation
or telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \todo{citations}. More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
-- formalized in ecological terms as the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} -- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behavior is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin_quantitative_1998}.

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states stays
relatively constant; changes in behavioural mode switching based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred state and environmental
conditions \citep{fryxell_multiple_2008}. \todo{I \emph{think} Fryxell
  et al use ad hoc rather than formal methods for their environment
  comparison? check this ...}  \cite{schliehe-diecks_application_2012}
consider changes in switching behaviour over time scales of six-hour
observation periods\todo{check?}, but
in general ecologists have used other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are independent,
\emph{finite mixture models} (FMM) -- a special case of HMMs where the
probability of state occupancy is independent of the previous state --
can adequately describe movement. \todo{are there animal movement
  studies that use FMMs?}  When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities changed systematically
\todo{smoothly? deterministically?} over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We take data from 
\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous
HMMs \todo{we will clarify below that they actually used HSMMs} 
to analyze hourly movement data
from 18 Florida panthers (\emph{Puma concolor coryi}).

\cite{van_de_kerk_hidden_2014} found that the best-fitting HMMs
according to BIC incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,van_de_kerk_hidden_2014}.
As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{van_de_kerk_hidden_2014}'s model-based approach to
identifying the number of behavioural modes.

@@

\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.  
\end{enumerate}
The likelihood above is the likelihood of the observed sequence given a particular hidden sequence. In order to calculate the likelihood of the observed sequence given the HMM, we need to sum over all possible hidden sequences. This direct method is computationally inefficient, but there exist more efficient ways to computation the likelihood and estimate parameters numerically \citep{zucchini_hidden_2009}.

>>>>

\documentclass{article}
\usepackage{natbib}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout] %{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\bibliographystyle{chicago}
\title{ HMM}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

\keywords{Hidden Markov, temporal heterogeneity}

@@

Given a sequence of animal movements observed via direct observation
or telemetry, movement models aim to find a parsimonious description
that can be used to understand past movements as well as predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \citep{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\citep{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012}
-- formalized in ecological terms as the ``multiphasic movement
framework'' \citep{fryxell_multiple_2008} -- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behavior is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\citep{okubo_diffusion_1980,turchin_quantitative_1998}.

@@

HMMs have typically been used to model movements over short time
scales, where the probability of switching between movement states stays
relatively constant; changes in behavioural mode switching based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred state and environmental
conditions \citep{fryxell_multiple_2008}. \todo{I \emph{think} Fryxell
  et al use ad hoc rather than formal methods for their environment
  comparison? check this ...}  \cite{schliehe-diecks_application_2012}
consider changes in switching behaviour over time scales of six-hour
observation periods, but
in general ecologists have used other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are independent,
\emph{finite mixture models} (FMM) -- a special case of HMMs where the
probability of state occupancy is independent of the previous state --
can adequately describe movement. \todo{are there animal movement
  studies that use FMMs?} 
%% \citep{tracey_mapping_2012}  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities changed systematically
\todo{smoothly? deterministically?} over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or switching
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying switching probabilities -- in particular,
switching probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We take data from 
\cite{van_de_kerk_hidden_2014}, who used temporally homogeneous hidden semi-Markov
models, HSMMs (an extension of HMMs that allows flexible modelling of dwell time)

%%\todo{we will clarify below that they actually used HSMMs} 
%%to analyze hourly movement data
%%from 18 Florida panthers (\emph{Puma concolor coryi}).

\cite{van_de_kerk_hidden_2014} found that the best-fitting HSMMs
according to BIC incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
van_de_kerk_hidden_2014}.As \citeauthor{van_de_kerk_hidden_2014} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{van_de_kerk_hidden_2014}'s model-based approach to
identifying the number of behavioural modes.

@@

\item Emission Probability: $P($\boldmath{$y_{k}$}\unboldmath $\mid z_{k},$\boldmath{$x_{k}$}\unboldmath$)$, a vector of observation density \boldmath{$y_{k}$} \unboldmath condition on covariates \boldmath{$x_{k}$} \unboldmath at state $z_{k}$ at time $t=k$.  
\end{enumerate}
The likelihood above is the likelihood of the observed sequence given a particular
hidden sequence. In order to calculate the likelihood of the observed sequence given the
HMM, we need to sum over all possible hidden sequences. This direct method is computati
nally inefficient, but there exist more efficient ways to computation the likelihood and
estimate parameters numerically \citep{zucchini_hidden_2009}.
